%!TEX root = edl.tex



\section{Entailment}

Another problematic feature of \lone\ (somewhat related to the foregoing by the deduction theorem) are the so-called \emph{paradoxes of entailment}. These are the following valid sequents: \begin{itemize}
	\item $\phi, \neg\phi \vDash \psi$;
	\item $\phi \vDash \psi \vee \neg \psi$.
\end{itemize}
But, some have objected, these are terrible inferences – we should never conclude, on the basis of contradictory premises, anything; and while it may be safe to conclude a tautology from any premises, that is hardly a good argument to persuade someone that the conclusion is a tautology. 

\paragraph{Relevance} Critics of these sequents have maintained that the failure here is one of \emph{relevance}. The arguments may be valid in the narrow sense, because there is no structure which satisfies the premises without also satisfying the conclusion (in the former case, because no structure satisfies the premises; in the latter, because every structure satisfies the conclusion). But validity in this narrow sense is too narrow – what we ordinarily understand to be a successful valid argument excludes these trivial cases of validity. The conclusion in a \emph{genuinely} valid argument – these critics say – should follow from, and be relevant to, the premises.

\paragraph{The Lewis Argument} Yet there is a very simple argument, using ordinary English reasoning which is apparently impeccable, to support these `problematic' sequents. It is known as the \emph{Lewis argument}, after C. I. Lewis, one of its modern rediscoverers. 
\begin{quote}
	\begin{exe}
	\ex $\phi$ \hfill Assumption
	\ex $\phi$ or $\psi$ \hfill Disjunction introduction, 1
	\ex $\neg \phi$ \hfill Assumption 
	\ex $\psi$ \hfill  Disjunctive syllogism, 2, 3
\end{exe}
\end{quote}
`Disjunctive syllogism' is the rule that from $\phi \vee \psi$ and $\neg \phi$ one can infer $\psi$. (A problem asks you to show that this is valid derived rule in the system \lone; it is obviously a good rule for the English `or'.) 

This argument seems to show that, even ordinarily, this form of apparently irrelevant argument is valid. And who could object to disjunction introduction or disjunctive syllogism, which seem like about the most basic things one can safely say about disjunction.

\paragraph{The relevantist response – reject Disjunctive Syllogism} The relevantist has a response to this argument – reject disjunctive syllogism. For, they say, it goes wrong in precisely this circumstance. The inference works normally because, given a true disjunction with one false disjunct, we can infer to the truth of the other disjunct \emph{without} checking to see whether it is true. (Disjunctive syllogism was called `the dog' by its Stoic inventors, because `even a dog uses this form of inference when it comes to a fork in the road, sniffs down one branch, and not finding the scent there immediately takes off down the other branch, without stopping to sniff' \citep[99--100]{burphilo}.) But in this case, $\phi$  is also an assumption – so the fact that $\phi \vee \psi$ has a disjunct whose denial is an assumption doesn't entail $\psi$. If there are cases where we can have inconsistent assumptions, like this, we should expect disjunctive syllogism to go wrong.

\paragraph{Dialethism} Some truly radical support for this line of argument can be found from the \emph{dialethist} position that some sentences are \emph{both} true and false. If $\phi$ is one such sentence, then both assumptions can be true ($\phi$ because it is true, $\neg\phi$ because $\phi$ is false). But $\psi$, since it is arbitrary, can be chosen to be a plain falsehood, giving a situation  which is a counterexample to disjunctive syllogism. Dialethists motivate their non-standard semantics by appeal to particular problem cases that don't seem to admit of any other solution. The best known is \begin{exe}
	\exi{L} L is false.
\end{exe} The sentence L displayed above says of itself that it is false. If it is true, then L is false (as that is what it says). But if it is false, then what it says is false, so it must be true. So L is true iff it is false. The dialethist takes this proof at face value – L is both true and false. But, despite the great difficulties more standard approaches to the Liar paradox L have, few have found dialethism compelling (though it has surprisingly able defenders). 

 \paragraph{Inconsistent belief} Dialethism is too incredible to convincingly motivate the rejection of disjunctive syllogism. A weaker motivation is found in reasoning about what we believe. If you are like me, you probably have some inconsistent beliefs that you are not aware of. It is absurd to say that I therefore believe everything, just because I have beliefs that entail everything. So it seems, the set of my beliefs is not plausibly closed under classical entailment (that is, the set of my beliefs does not contain everything which is  entailed by some things in my set of beliefs). Disjunctive syllogism seems to fail for my beliefs; just because $\phi\vee\psi$ and $\neg\phi$ are among my beliefs does not mean that $\psi$ is one of my beliefs; for $\phi$ may be one of my beliefs as well. The logic of belief, it may be said, is relevant. (The original application of this kind of idea was to simple inference-drawing computers – if they are not aware of when the body of information input contains a contradiction, you should not allow them to draw conclusions using full classical logic.) The problem with all this is that it doesn't really seem like logic any more. There are no counterexamples to disjunctive syllogism, that is, structures where $\phi\vee\psi$ and $\neg\phi$ are true but $\psi$ is false. Rather, there are cases where \emph{according to my beliefs} $\phi \vee \psi$ and $\neg\phi$ are true, and also \emph{according to my beliefs} $\psi$ is not true. But it is already well-known that, when we have a non-truth-functional operator like `according to my beliefs', it won't generally be true that classically valid sequents remain correct when each sentence in the sequent is in the scope of the operator. (Think of the non-truth-functional operator `possibly': $\phi, \psi \vDash \phi \wedge \psi$ is correct, `possibly $\phi$, possibly $\psi \vDash$ possibly $\phi \wedge \psi$' is not correct – think of the case where $\psi$ = `the coin toss lands heads', $\phi$ = `the coin toss lands tails'.) 



\section{Designators}
\paragraph{Designators in \ltwo\ and English}

The formal properties of \ltwo\ have been established; now we begin a relatively informal investigation into the possible connections between the semantics of \ltwo\ and the semantics of English.
We begin in this section with a discussion of \emph{designation} in both languages.

\paragraph{Direct Reference}
\ltwo\ embodies a particularly simple view of the meaning of a constant.
 \begin{definition}[Direct Reference]
 	A referring expression \emph{directly refers} iff the meaning of the name is just its referent, the object denoted by the expression.
 \end{definition}	
Constants in \ltwo\ directly refer, because the meaning (semantic value) of $a$ in $\mathscr{A}$ is $\val{a}{A}$, which is just an item in the domain of $\mathscr{A}$.


This is to be contrasted with any thesis of \emph{indirect} reference, according to which the semantic value of a denoting expression is something other than the referent. The primary example here are definite descriptions; if Russell's account (recall chapter \ref{c:lequals}) is correct, for example,  the meaning of a description is somehow to be extracted from a complex quantified sentence; the referent (if any) is determined by the meaning, but is not the meaning. This can clearly be seen if we accept Russell's treatment of ‘The present king of France is bald’, where the sentence is meaningful even though the definite description fails to denote anything. Given that the meaning of the sentence is determined by the meanings of its constituents, the definite description noun phrase must have a meaning, even though it lacks a referent, and hence cannot be a directly referential expression.



\paragraph{Rigid Designation} Direct reference is related to, but distinct from, the following notion:
\begin{definition}[Rigid Designator]
	A referring expression $a$ is a \emph{rigid designator} iff the referent of $a$ is the same object in every possible situation (possible world) in which that object exists.\footnote{Be careful not to confuse this with the claim that every possible use of ‘$a$’ refers to the same thing – this would be false. The idea is rather, when we actually evaluate the possibility of various things about $a$ in some possible world $w$, we continue to hold fixed the actual meaning of $a$, rather than (for example) what the residents of $w$ themselves would mean by their own utterance of ‘$a$’. Suppose we consider the possibility in which Aristotle hadn't been called ’Aristotle’: we are still talking about the person actually called ‘Aristotle’, even though residents of that possibility do not use that name to denote anyone.}
\end{definition}
It may seem that this doesn't hold in \ltwo, because if $\mathscr{A}\neq\mathscr{B}$, then it may be that  $I_{\mathscr{A}}(a)\neq I_{\mathscr{B}}(a)$. But for all that it may hold; the crucial fact to recognise is whether, \emph{in a given situation}, when we consider what is possible for $a$, we consider what is possible for what $a$ refers to – in every situation in which that thing exists. Since \ltwo\ has no resources for discussing possibility, it is a moot point whether or not rigid designation holds for it. 

If an expression is directly referential, in a language with resources for discussing possibility and necessity, then it looks like it will also be a rigid designator. Assume that we evaluate ‘Possibly, $F(a)$’ by seeing whether there is a possible world in which $F(a)$. If $a$ directly refers, the evaluation of ‘Possibly $F(a)$’ will proceed by evaluating whether actual meaning of $a$ – which is just some object $\mathbf{a}$, given direct reference – satisfies $F$ at each possibility in which $\mathbf{a}$ exists. So the actual meaning of $a$, evaluated at each world, will remain constant.

The classic example of a non-rigid designator is also a simple definite description, like ‘the tallest person’. Suppose Jacques is the tallest person. Still, I can perfectly well say ‘Possibly, Gill is the tallest person’, which is \emph{not} synonymous with ‘Possibly, Gill is Jacques’. At the possible world in which Gill is the tallest person, the expression ‘the tallest person’, with its actual meaning, denotes her, though it doesn't denote her actually. 

Things are complicated by the existence of \emph{rigidified descriptions}. As the name suggests, they are rigid designators, but they are also descriptions which refer indirectly. Consider ‘the actual tallest person’: intuitively, the meaning of this description evaluated at every possibility determines it to have a constant referent, of Jacques. Because of this, ‘Possibly, Gill is the actual tallest person’ does seem to entail that possibly, Gill is Jacques. 
	


\paragraph{Names and Direct Reference in English}
For names in English, like `Antony Eagle' or `Aristotle', there is no straightforward semantics for names as there is for constants in \ltwo.

On the one hand, there is considerable syntactic and semantic evidence that names in English function rather like the constants of \ltwo. Consider \begin{exe}
	\ex Antony Eagle is lecturing now.\label{s}
\end{exe}  (\ref{s}) is true  iff the thing \textbf{Antony Eagle}, designated by `Antony Eagle', has the property \textbf{lecturing} denoted by `is lecturing', \textbf{now}, i.e., at the time designated by (this use of) `now'. These seem remarkably like the clauses for the satisfaction of atomic formulae of \ltwo; in that sense it looks like the meaning of `Antony Eagle' is just the thing it refers to.

\paragraph{Objection to Direct Reference: Informative Identities}
 If `Hesperus' denotes Hesperus, and `Phosphorus' denotes Phosphorus, then, since Hesperus \emph{is} Phosphorus (both are names for Venus), then the meanings of `Hesperus' and `Phosphorus' are the same. So then why aren't these sentences synonymous?: \begin{exe}
		\ex Hesperus is Hesperus.
		\ex Hesperus is Phosphorus.
	\end{exe}



This objection trades on the fact that we receive information when told the second claim, but we do not when told the first; hence they must yield different information, and not be synonymous after all. One possible response is to say that \emph{for all we know}, we are in a situation where `Phosphorus' names something other than Venus; and what we are told by the second sentence is a contingent fact about English, namely, that `Phosphorus' means Phosphorus!

\paragraph{Objection to Direct Reference: Empty Names}
If the meaning of a name is its referent, then `Santa Claus is jolly' is meaningless, yet we judge it to be true. And don't say it is true `in a fiction' – for fictional English is the same as non-fictional English, so if it is meaningless outside of the fiction, it is just as meaningless inside. But set that aside: \emph{negated existentials} show that `Santa Claus' is meaningful, because `Santa Claus does not exist' is straightforwardly true.



To be more explicit: `Santa Claus does not exist' apparently expresses a truth. This can be rendered in the language \lequ\ by this translation: `$\neg \exists x (x= a)$', where `$a$' denotes Santa Claus. But since `$a=a$' expresses a logical truth, it follows that `$\exists x (x = a)$' expresses a logical truth of \lequ, denying the apparent truth which we translated. But this looks like a problem in English, as well as in formal languages: since `Santa Claus is identical to Santa Claus' also expresses a truth, the English claim `Something is identical to Santa Claus' is also true. But if something is identical to Santa Claus, Santa Claus does exist. To deny this is to deny claims that seem tautologous in English. 

This is a very serious objection to direct reference theories. Many have chosen to bite the bullet, and accept that such sentences are in fact meaningless. Others have chosen to accept some secondary notion of meaning, like a canonical description associated with the empty name, to play a name-like role in these kind of claims. But there is no consensus on how to deal with this problem.

\paragraph{Objection to Direct Reference: The Role of Descriptions}
Many names are simply introduced by descriptions; and certainly what most of us aim to communicate when we use a name is a certain individual satisfying a certain canonical description that the hearer is familiar with – if there were no such description, `how do people ever use names to refer to things at all?' \citep[28]{krinamne}. 
\begin{description}
	\item [Reference Fixing] Some descriptions are used to establish the reference: `Julius' may be introduced into the language as `the inventor of the zipper'. But it is no part of the meaning: consider the counterfactual situation where Julius gets pipped by his rival.
	\item [Background knowledge] If one knows a lot about Julius, and one's hearers do too, then one will expect that background knowledge to be useable when one uses the name `Julius'. It may be difficult if not impossible for a speaker themselves to separate the meaning of `Julius' from what they know about Julius.
\end{description}

\paragraph{Opaque Contexts}
 Consider \begin{exe}
	\ex Lois Lane believes that Superman is Superman;
	\ex Lois Lane believes that Superman is Clark Kent.
\end{exe}
It is fairly clear that in the context `Lois Lane believes that Superman is $x$', one cannot straightforwardly substitute co-referring expressions into $x$ and get a true sentence. Such \emph{opaque contexts} – those which do not permit substitution \emph{salva veritate} – are a problem for the direct reference theory, of course, but also for other theories that think the objects of belief are propositions concerning the entities one has the belief about. 

Obviously \ltwo\ lacks opaque contexts, as theorems about substitution of co-referring constants proved in Chapter \ref{c:l2} show. This is an expressive weakness of \ltwo\ compared with English.



\paragraph{Other Designators in English}
There are several other natural language expressions other than names and definite descriptions which have a designator role in English. \begin{description}
	\item [Indexicals] such as `I', `you', `here', `now', `actually'. This class may also include \emph{demonstratives} like `this', `that'.
	\item [Count Nouns] such as `chair', `boy', `piece of furniture'.
	\item [Non-Count Nouns] such as `gold', `dirt', `intelligence'.
\end{description}
Count nouns can be handled by the apparatus of quantification in \ltwo, rather than as designating expressions directly. 

\paragraph{Indexicals}
The characteristic of an indexical expression is \emph{context sensitivity}. `I' refers to different people when uttered by different people: to me when I say `I am tired', to you if you utter that same string of words. Similarly with other expressions in this category, as they are sensitive to features of context – where the context of an utterance includes the speaker, time, place, etc., of the utterance – and thus may have different referents in different contexts.

Despite the context-sensitivity of indexical \emph{reference}, their meaning appears constant. Distinguish (following Kaplan):  \begin{description}
	\item [Content] The content of an expression is its semantic value;
	\item [Character] The character of an expression is a function from context to content.
\end{description} The thesis is that indexicals have a constant character, but a variable content; and character is plausibly a kind of meaning.

Pronouns – `she', `they', `it' – may perhaps be analysed similarly.




\paragraph{Non-Count Nouns}
Non-count nouns are a certain type of \emph{non-plural} referring expression. One reasonable test is whether `$\mu$'  can be grammatically substituted into the expression `I want some $\mu$'. 
 These fall into at least two kinds: \begin{description}
	\item [Mass] Mass nouns, like `gold' or `dirt', denote a kind of stuff or substance, rather than a particular thing. They typically display \emph{cumulative reference}: if `being gold' is true of $a$ and $b$, then it is true of the \emph{sum} $a+b$ which has just $a$ and $b$ as parts.
	\item [Abstract] Abstract nouns, like `intelligence' or `generosity', which seem to denote \emph{qualities} that something may possess, but can themselves have things predicated of them.  
\end{description}
A full treatment of these nouns cannot be carried out in \ltwo; constants referring to qualities, that we naturally translate as predicates like `being gold', may require a \emph{second order} logic, which permits quantification over predicate variables as well as individual variables. 







\section{More on Relations}
 
 \paragraph{Expressing Properties of Binary Relations in \ltwo}
 
We talked about the properties of binary relations, modelled as sets, back in \autoref{c:sets}.

 We can use formulae of \ltwo\ to express various properties of relations on an entire domain. For example \begin{description}
 	\item [Reflexivity] A relation $R$ is reflexive on $D_{\mathscr{A}}$ if $R = \val{P^{2}}{A}$ and\\ $\val{\forall x P^{2}xx}{A}=T$. 
\item [Transitivity] A relation $R$ is transitive on $D_{\mathscr{A}}$ if $R = \val{P^{2}}{A}$ and $$\val{\forall x \forall y \forall z (P^{2}xy \wedge P^{2}yz) \to P^{2}(xz)}{A}=T.$$ 
\item [Symmetry] A relation $R$ is symmetric on $D_{\mathscr{A}}$ if $R = \val{P^{2}}{A}$ and\\ $\val{\forall x \forall y  (P^{2}xy \to P^{2}yz}{A}=T$. 
 \end{description}

\paragraph{Relations in English and \ltwo: Intension and Extension}

Is the semantic value of an English predicate expression a property or relation as those are interpreted in \ltwo-structures?

The property `is a person' on the domain of Oxford students is the same property as `is a student'; but even though they coincide in their members, we do not think these predicates are synonymous. We may distinguish the property given \begin{itemize}\item \emph{in extension}, by the list of things which satisfy it; or \item \emph{in intension}, by a rule.\end{itemize} On this domain, two different rules give the same extension; and intuitively in English the meaning is the property given intensionally.

\paragraph{Relations in English and \ltwo: Sparse and Abundant Properties}


 \emph{Any} set of pairs is a binary relation. Yet this means that `is identical to' is just as legitimate a relation as `is within 21 metres of'; whereas we think the former is a more fundamental and non-accidental relation than the latter. We may think that the `real' properties are sparse and fundamental, and the abundant predicates like `grue' and `within 21 metres of' are less so, and do not correspond to genuine properties – but \ltwo\ hardly allows us to draw that distinction.

So despite the fact that `grue' and `green' are both meaningful predicates of English, we want to give quite different things to be their semantic values – perhaps just a set of things for `grue', but a real property \textbf{greenness} to be the semantic value of `green'.

Despite these problems, the case against English having a similar semantics for predicates to \ltwo\ is rather weaker than that against English having a similar semantics for designators.


{\small
\subsection*{Further Reading}
\addcontentsline{toc}{subsection}{Further Reading}




 The best defence of dialethism is \citet{priinco}; the application of relevant logic to inconsistent data sets is \citet{belusefov}.  

A wonderful book on names and description is \citet{krinamne}. A modern classic: amazingly, it is more or less a transcription of a series of lectures originally delivered without notes. A useful source on empty names is \citet{capempna}. The topic of empty names is one main topic of another famous series of lectures by \citet{krirefex}.



\subsection*{Exercises}
\addcontentsline{toc}{subsection}{Exercises}


\begin{enumerate}
	
	\item 	\begin{enumerate} \item Show that the following derivation rules are acceptable in our framework: \begin{enumerate}
		\item \emph{Modus Ponens}: If $\vDash \phi$ and $\vDash \phi \to \psi$, then $\vDash \psi$.
				\item \emph{Disjunctive Syllogism}: If $\vDash \phi \vee \psi$ and $\vDash \neg \phi$, then $\vDash \psi$.
				\item \emph{Reductio ad Absurdum}: If $\phi \vDash \psi$ and $\phi \vDash \neg \psi$ then $\vDash \neg \phi$. \end{enumerate}
	\item
		 In our system each acceptable rule of inference has a corresponding correct semantic sequent (for example, the sequent $\phi, (\phi \to \psi) \vDash \psi$ corresponds to  \emph{modus ponens}). Consider now this rule of inference in English: if $\phi$ is a truth of logic, then `Necessarily, $\phi$' is too.  Is this rule intuitively correct? Is the corresponding sequent `$\phi \vDash_{\text{English}} \text{Necessarily } \phi$' intuitively correct?
		\end{enumerate}


\item Say that $\phi \vDash \psi$ is a \emph{perfect} sequent iff it is correct, and $\phi$ is satisfiable, and	$\psi$ is not a tautology. Say that a sequent is \emph{perfectible} iff it is a substitution instance of a perfect sequent. \begin{enumerate}
	\item Show that there must be a sentence letter in common between $\phi$ and $\psi$ is a perfectible sequent. (Hint: use the Craig Interpolation Theorem (page \pageref{thmcraig}), plus properties of substitution.)
	\item Give examples to show that perfectible entailment is not transitive (i.e., that there are cases where $\phi$ perfectibly entails $\psi$, where $\psi$ perfectibly entails $\chi$, but where $\phi$ does not perfectibly entail $\chi$.)
	\item Which of the structural rules of \S\ref{twostruct} (page \pageref{twostruct}) are perfectibly inappropriate (i.e., take perfectible sequents to non-perfectible sequents.)?
\end{enumerate}
	
	
\item Which of the relations expressed by the following English predicates are equivalence relations: \begin{enumerate}
	\item `$x$ entails $y$', on the domain of sentences of \ltwo.
	\item `$x$ is logically equivalent to $y$', on the domain of sentences of \ltwo.
	\item `If $\{x\}$ is consistent, then $\{x\}\cup \{y\}$ is consistent' on the domain of sentences of \ltwo.
\end{enumerate}

\item  Show that $\forall x \forall y (\neg x=y\to (Pxy \vee Pyx) \equiv \forall x \forall y (x=y \vee Pxy \vee Pyx)$. 
	
\item Show by suitable reasoning that
    in a finite domain, for any \emph{partial order} $R$,  $\exists x \forall y (Ryx \to
    x=y)$. Give a counterexample to this condition in an infinite domain. 

\item Give a graph, on some non-empty domain, of a relation $R$
	which satisfies this condition:
	$\forall x \forall y (Rxy \to \exists z ((x \neq z) \wedge (y \neq
	z) \wedge Rxz))$. Can you give an example of a relation which
	satisfies this condition (being sure to specify the domain)?
	
	







\item	We might normally expect `is similar to' to be a symmetric relation: after all, if there is a respect in which $a$ is similar to $b$, then $b$ must be similar to $a$ in that very same respect. But many people seem to judge that similarity is \emph{not} symmmetric: \begin{quote}
		When people are asked to make comparisons between a highly familiar object and a less familiar one, their responses reveal a systematic asymmetry: The unfamiliar object is judged as more similar to the familiar one than vice versa. For example, people who know more about the USA than about Mexico judge Mexico to be more similar to the USA than the USA is to Mexico. \citep[520]{kunsocco}
	\end{quote} (You might think also of the fact that it is much more natural to say that children resemble their parents, than that parents resemble their children.)

	Can you provide a rationale for these psychological results? Do they indicate that people are systematically mistaken about the meaning of the relational predicate `is similar to', or do they indicate that our theory of similarity in terms of matching respects of similarity is incorrect?

\end{enumerate}

}



	








