%!TEX root = edl.tex

\section{Sets}

A \emph{set} is a collection of objects. Like a collection, it is a single thing, not a plurality. If $X$ is a set, the objects in $X$ are called its \emph{elements} or \emph{members}. We write $x \in X$ for `$x$ is an element of $X$' (and $y \notin X$ for `$y$ is not an element of $X$').

 Sets are \emph{individuated} by their members; or as we sometimes say, sets are individuated purely \emph{extensionally}. There must be something to a set over and above its members – otherwise it would be simply a plurality identical to those members. But whatever that ‘something’ is, it is minimal enough that there cannot be two distinct sets which share the same members.Therefore, if $X$ and $Y$ name sets which have all and only the same members, $X = Y$. So we can specify a set by listing the members, conventionally in curly braces: $X = \{x_{1},...,x_{n},…\}$. 

 In this, sets differ from \emph{sequences}: two distinct sequences can have the same members as long as they occur in a different order in the two sequences. Some say that sets also differ from wholes in this respect: while you can make only one set from a given collection of members, you can make more than one object from a given collection of parts, if you arrange those parts in different ways. (This last example is however controversial: some think that appearances here are deceptive.) 

A set is not \emph{composed} of its members in any straightforward sense: it is not a whole with its members as parts. We can see this most easily in the case of a \emph{singleton} set, a set with just one member. Clearly $x \in \{x\}$; but since $x$ itself need not be a set, $x \notin x$. Since sets are individuated by their members, and since $\{x\}$ has a member that $x$ does not, $x\neq\{x\}$. So a set is not to be identified with the fusion of its members, even when that latter object exists.

\paragraph{Axioms} The standard approach to set theory is to take this informal conception of a set as a collection, and set down \emph{axioms} that characterise the behaviour of sets. We've already seen one of the standard axioms of so-called Zermelo-Fraenkel set theory: the axiom of \emph{extensionality}: \begin{axiom}[Extensionality]
  For any sets $X$ and $Y$, if for all objects $x$, $x\in X$ iff $x\in Y$, then $X=Y$.
\end{axiom} I will below mention various axioms, the basic starting points of modern set theory, but we will largely proceed in an informal way. In particular, I will not be offering proofs from the axioms of all the set theoretic principles I state below.

\paragraph{Sets in Sets} One thing is worth being explicit about at the outset. \emph{Sets can have other sets as members.} We might be used to thinking of collections as collections of material objects (rocks, china plates, etc.) But since a set is not identical to its members, and we can talk about and quantify over sets, they are also objects available to be the members of further sets. We might \emph{start} with the non-sets, or \emph{ur-elements} as they are sometimes known; but once we make sets of them, we can then make further sets which include sets as members, and then further sets involving sets of sets, etc. This is the so-called \emph{iterative conception of a set} \citep{boolos}.


\paragraph{Defining Sets}

We've seen one way to define a set: list the members. But what if there are too many to list? 

One thing we can do is define sets \emph{inductively}. This means: given a base case, and a \emph{generating relation}, we define a set that includes the base case, and anything related to a member of the set by the generating relation, and   nothing else (the \emph{closure condition}). Here's an example. \begin{definition}[Your Ancestors] 
   \emph{Base case:} Your mother and father are ancestors.\\ \emph{Generating Relation:} Any parent of an ancestor is an ancestor.\\ \emph{Closure Condition:} Nothing else is an ancestor of you.
 \end{definition}

Let $A_{ae}$ be the set of my ancestors. Any member of $A_{ae}$ is a parent of a parent of\ldots a parent of me, with $n >0$ occurences of `a parent of'. The property `is an ancestor of AE' is called the \emph{ancestral} of the property `is a parent of AE'. And more generally we can say that the relation $R$ `$x$ is an ancestor of $y$' is the ancestral of the relation `$x$ is a parent of $y$'. Note that the set $A_{ae} = \{x: R(x,ae)\}$

Alternatively, we can specify a set by stating a condition $F$ that the members all satisfy, typically in some formal mathematical language. Then we can specify the set of things $x$ such that each $x$ is $F$. We write this $\{x: F(x)\}$. If $Y = \{x:F(x)\}$, then $y \in Y$ iff $F(y)$. But in fact any collection of items forms a set: there needn't be a natural condition all the members satisfy.

 Not just any condition will define a set. Consider the condition $F = x \notin x$. This defines a set $\mathbf{R} = \{x: x\notin x\}$, the set of all non-self-membered things.

 Is $\mathbf{R}$ a member of $\mathbf{R}$? Assume $\mathbf{R} \in \mathbf{R}$. Then $\mathbf{R}$ must meet the defining condition $F$, so that $\mathbf{R} \notin \mathbf{R}$. But since $\mathbf{R}$ cannot both be and not be a member of itself, the assumption must be wrong. So $\mathbf{R}\notin \mathbf{R}$. But then $\mathbf{R}$ does meet $F$, and therefore $\mathbf{R} \in \mathbf{R}$. A contradiction follows. This is known as \emph{Russell's paradox}\label{russ}.

The problem is assuming that \emph{any} condition at all defines a set. The alternative is simple: to assume that, \emph{if we are given a set $X$}, then any condition at all will \emph{separate} the members of $X$ into those which meet the condition and those which do not. That is: for any set $X$, and any condition $F$, there will exist a set $Y$ which contains everything which is both a member of $X$ and satisfies $F$. This is the axiom of \emph{separation}, so called because it separates out those members of a set which have a certain specified feature.\begin{axiom}[Separation]
	If $X$ is a set that is already given, and $F$ is any condition, then there exists a set $Y$ consisting of all the members of $X$ which meet the condition $F$. (This may be trivial, in that all or no members of $X$ meet the condition.) That is, for any $X$ and $F$, this specification is well-defined and guaranteed to denote a set: $\{x\in X: F(x)\}$.
\end{axiom}


\paragraph{Relations Between Sets} Some important relations of inclusion and exclusion between sets can be defined, once we are assured that some sets exist.
\begin{definition}[Subset] A set $X$ is a \emph{subset} of $Y$, written
   $X \subseteq Y$ if every
    member of $X$ is also a member of $Y$: for every $x$, if $x \in X$
    then $x \in Y$. $X$ is a \emph{strict subset} (or \emph{proper
    subset}) of $Y$ if $X
    \subseteq Y$ and $Y \not\subseteq X$ (written `$X \subset
Y$').\end{definition}
Trivially, for all sets $X$, $X\subseteq X$. It also follows that if $X \subseteq Y$ and $Y \subseteq X$ then $X=Y$ (easy). And also if $X \subseteq Y$ and $Y \subseteq Z$ then $X \subseteq Z$.
Note that if $F$ is some condition, the set $\{x:x\in X \text{ and } F(x)\}$, which is guaranteed to exist by the axiom of Separation, is always a subset of $X$.
\begin{definition}[Superset] A set $X$ is a \emph{superset} of $Y$, written   $X \supseteq Y$ if every    member of $Y$ is also a member of $X$: for every $x$, if $x \in Y$    then $x \in X$. $X$ is a \emph{strict superset} (or \emph{proper superset}) of $Y$ if $X    \supseteq Y$ and $Y \not\supseteq X$ (written `$X \supset Y$').\end{definition}
\begin{definition}[Disjoint] Sets $X$ and $Y$ are \emph{disjoint} iff there is no $x$ which is a member of both.	
\end{definition}


\paragraph{Operations on Sets} The foregoing relations are defined between existing sets. But there are also generating operations, that given some existing sets yield some new sets. For each of these generating operations, we need to give a definition of the properties of the operation. But we also need an axiom to ensure that that operation succeeds, that is, that there is a set which satisfies the definition we have set down. The axiom of Separation supports our first two definitions: 
\begin{definition}[Intersection] If $X$ and $Y$ are sets, the set $X
    \cap Y$ which contains only members of both is called the {\em
    intersection.} 

If $\mathbf{X}=X_{1},\ldots,X_{n},\ldots$ is a set of sets, $\bigcap \mathbf{X} = X_{1} \cap \ldots \cap X_{n} \cap \ldots$.
\end{definition}
 \begin{definition}[Relative Complement] If $X$ and $Y$
	are sets, then $X \setminus Y$ is the set of all members of $X$ which
	are \emph{not} members of $Y$, called the \emph{relative complement} of
	$Y$ with respect to $X$. \end{definition}
Because both of these definitions involve taking a set that is already given and constructing a subset, the existence of the resulting sets is supported by the axiom of Separation. In the case of intersection, if $X$ and $Y$ exist, $X \cap Y$ exists: $X\cap Y = \{x \in X:x\in Y\}$. Alternatively, we can denote this set as the set of any $x$ that meets the conjunctive condition of being in both $X$ and $Y$: $\{x: x\in X \text{ and } x \in Y\}$. In the case of relatively complement, if $X$ and $Y$ exist, $X\setminus Y$ exists: $X \setminus Y = \{x \in Y: x\notin Y\}$.

What if we want to make a set which isn't a subset of an already given set?
\begin{definition}[Union] If $X$ and $Y$ are sets, then $X \cup Y$
    (read `the \emph{union} of $X$ and $Y$') is the set which contains all the members of $X$ and all the members of $Y$. 

If $\mathbf{X}=X_{1},\ldots,X_{n},\ldots$ is a set of sets, $\bigcup \mathbf{X} = X_{1} \cup \ldots \cup X_{n} \cup \ldots$.
\end{definition}
The existence of unions is supported by this axiom: \begin{axiom}[Unions]
  Suppose $\mathbf{X}$ is any set of sets (i.e., a set with only other sets as its members). For any such $\mathbf{X}$, there is a set $Y$ such that, for any $z$, $z\in Y$ iff $z$ is a member of one of the members of $\mathbf{X}$.  In short: if $X$ and $Y$ exist, $X\cup Y$ exists: $X \cup Y = \{x:x\in X \text{ or } x \in Y\}$.  
\end{axiom} That is, $Y$ is the set containing all the things which are in any of the sets in $\mathbf{X}$. If $X$ and $Y$ are sets, there is a set $\{A,B\}$.\footnote{This is in fact another axiom, the axiom of Pairing.} It then follows from the axiom of Unions that there is a set which contains all the members of $X$ and all the members of $Y$. By extensionality, there is just one such set; we call it $A \cup B = \bigcup\{A,B\}$.

Note the links  between $\cap$ and `and', and $\cup$ and `or'.

\paragraph{The Empty Set}

If $X$ is a set, then by the definition of relative complement, $X\setminus X$ is a set. Suppose $X\setminus X$ has a member, $z$.  Then 	$z\in X\setminus X$ iff $z \notin X \text{ and } z \in X$. Contradiction; so $X\setminus X$ has no members. This set is the \emph{empty set}, written $\emptyset$. $Y\setminus Y$ is also empty; by extensionality, $X\setminus X=\emptyset=Y\setminus Y$; there is only one empty set.

The empty set is the collection which has nothing in it. It corresponds to the empty list. While the empty collection has no members, that doesn't mean that it itself is nothing: it is the unique thing which is a set and also has no members. 

 If $X$ is any set, then for any $x$, if $x \in \emptyset$ then $x\in X$ – for the empty set has no members at all. By the definition of subset, therefore, $\emptyset \subseteq X$, for any $X$.

 It is easy to show that when $X$ and $Y$ are disjoint, $X \cap Y = \emptyset$.

\paragraph{Absolute Complement and the Universal Set}

The empty set is the smallest possible set in terms of its members. Is there a largest set? If there is a \emph{universal set} $\Omega$ – a set which has everything (every individual \emph{and every set}) as a member – we could define something a bit closer to negation than relative complement. We could define: $-X = \Omega \setminus X$; it would be the set of all things which are not in $X$.

This is \emph{not} possible. Suppose there were a universal set. Then the axiom of separation, that if $X$ is a set then there is a set $Y$ which is a subset of $X$ and such that for all $x\in X$ if $F(x)$ then $x\in Y$, will entail a contradiction.\footnote{Hint for exercises: Apply the condition $x\notin x$ to the universal set $\Omega$ in the axiom of separation, and we will construct the Russell set $\mathbf{R}$ perfectly legitimately. Since the Russell set does not exist, something's gone wrong: the only candidate is the assumption that there is a set of everything.} So there is no universal set. (Philosophical puzzle: how do we quantify over – talk about – all sets, as we have been doing, if there is no collection big enough to contain them all? How indeed can we even talk about \emph{everything} if there are too many things to be collected together into a set?) 

\paragraph{Power Set}

So far you might think that sets are just collections of individuals (i.e., non-sets). (Well, we've already mentioned the axiom of Pairing, so you probably don't think that.) But set theory is much more general than that: anything at all, including other sets, can be gathered into a set (as long as we don't try to gather too many things).

One useful set of sets is the \emph{power set} of a set $X$. This is, intuitively, the set of all ways to choose some members from $X$. We can choose no members; we can choose all the members; we can choose some but not all. Each such way of choosing determines a subset of $X$, so the set of all ways to choose members from $X$ is exactly the set of all subsets of $X$:
 \begin{definition}[Power Set]Given a set $X$, the \emph{power set} of $X$, denoted $\wp(X)$, is defined: $\wp(X) = \{Y: Y \subseteq X\}$. (It is another axiom of set theory that for any set, its power set exists – the aptly named Power Set axiom.)\end{definition} 
Notice that, for any $X$, $\emptyset \in \wp(X)$; and $X \in \wp(X)$. The name `power set' comes from this fact: 
\begin{theorem} If $X$ has $n$ members, $\wp(X)$ has $2^{n}$ members.
	\begin{proof}
		A subset of $X$ is a set containing perhaps some but not necessarily all the members of $X$. We can specify a subset by saying, of each member of $X$, whether it is in the subset.  Suppose the members of $X$ are $x_{1},\ldots,x_{n}$. Then a finite binary sequence, consisting of $1$s and $0$s, of length $n$ suffices to determine a subset of $X$, by the following principle: let $x_{i}$ be a member of the subset associated with $s$ iff the $i$th place in the sequence $s$ is a $1$. By extensionality, each subset of $X$ corresponds to exactly one such sequence. So how many finite binary sequences of length $n$ are there? Each place $i$ in the sequence can be either $1$ or $0$, so there are $2^{n}$ such sequences. So there are $2^n$ distinct subsets of $X$, and hence $2^{n}$ members of $\wp(X)$. 
	\end{proof}
 \end{theorem}

\paragraph{Ordered Pairs and Sequences}

As mentioned above, sets are identical iff they have the same members. How those members happen to be listed is irrelevant: $\{x,y\}=\{y,x\}=\{x,x,y\}$. So if we want to capture the idea of an \emph{ordered sequence}, it cannot be a set – order doesn't matter for sets, and one can't have repeated elements in a set. Yet we can model, or represent, sequences just using sets.
\begin{definition}[Ordered Pair] An \emph{ordered pair} $\langle x, y \rangle$ is defined as the unordered set $\bigl\{x,\{x,y\}\bigr\}$.
\end{definition}
This is not to say that ordered pairs are and always have been special kinds of sets. The existence of multiple equally good models for how to represent ordered pairs set-theoretically undermines any claim that ordered pairs are identical to sets. Kuratowski's original definition of an ordered pair was this different proposal:
\begin{definition}[Kuratowski Ordered Pair] A \emph{Kuratowski ordered pair} $«x,y» \eqdf \{\{x\},\{x,y\}\}$.
\end{definition}
 If we take seriously the identification of ordered pairs with sets, we need to answer which of these definitions identifies the `right' sets. But any such answer would be arbitrary, because it would involve picking one equally good representation of ordered pairs over another. The choice to model ordered pairs as in the text, or by Kuratowski's alternative, is a choice of model – it has no significance for what is being modelled, namely, sequences of length $2$.

Why do I say this? Because we can show that both our definition and Kuratowski's manage to capture the behaviour of  ordered pairs. We do so by establishing the fundamental principle of the identity of ordered pairs follows from the definitions, namely, that ordered pairs are identical iff they have the same first member, and the same second member. What we show is that, if we define ordered pairs as above, they have all the characteristic features of ordered pairs.
\begin{theorem}[Criteria of Identity for Ordered pairs]\label{cridorpa}
$\langle x,y\rangle = \langle u,v\rangle$ iff $x=u$ and $y=v$.
\begin{proof}
	$\langle x,y\rangle = \langle u,v\rangle$ iff $\left\{x,\{x,y\}\right\} = \left\{u,\{u,v\}\right\}$ iff (i) $x=u$ and $\{x,y\} = \{u,v\}$ or (ii) $x=\{u,v\}$ and $\{x,y\}=u$. Since (ii) is not possible – it would entail that $x$ is a member of one of its own members – it must be that (i), which in turn holds iff $x=u$ and $y=v$ (we need extensionality again to show that if $\{x,y\}=\{x,v\}$ then $y=v$).
\end{proof}
\end{theorem}
In the proof of this theorem, we had to appeal to the fact that no set is a member of a member of itself. (In proving the parallel criteria of identity theorem for Kuratowski ordered pairs, we need not appeal to this fact, showing which I leave as an exercise.) This fact obviously holds on the iterative conception of a set. It tacitly relies on this characteristic axiom of Zermelo-Fraenkel set theory: \begin{axiom}[Foundation]
	 Every non-empty set $X$ contains a member $y$  such that $X$ is disjoint from $y$: $X \cap y = \emptyset$.
\end{axiom} (This axiom is also known as \emph{regularity}.) \begin{theorem}
	No set is a member of itself. \begin{proof}
		Let $X$ be an arbitrary set. Clearly $X \in \{X\}$, which latter set exists by Pairing. By Foundation, since $\{X\}$ is non-empty, it has a member $y$ such that $\{X\} \cap y = \emptyset$. Since the only member of $\{X\}$ is $X$, it must be that $\{X\} \cap X = \emptyset$. But if $X\in X$, then $X$ and $\{X\}$ do have a member in common – namely, $X$ – and their intersection isn't empty. So $X \notin X$.
	\end{proof} 
\end{theorem} This theorem gives another route to avoiding Russell's paradox from earlier (page \pageref{russ}). But Foundation also enables us to rule out longer `cycles' of inclusion, since it entails that there are no infinite descending chains of set membership: every set bottoms out eventually in some members of its members of its members… that have no members themselves.

We can extend the idea of an ordered pair to an ordered sequence of any number of elements, as follows: \begin{definition}[Ordered $n$-tuple]
Let the notation  $⟪x_{1},\ldots,x_{n}⟫$ denote an ordered $n$-tuple, for $n\geqslant 2$. We offer an inductive definition. \begin{itemize}
	\item  \emph{Base case.} $⟪ x_{1},x_{2}⟫ \eqdf \langle x_{1},x_{2}\rangle$.
	\item \emph{Inductive step.} $⟪x_{1},…,x_{n+1}⟫ \eqdf \langle⟪ x_{1},…,x_{n}⟫,x_{n+1}\rangle$.
\end{itemize}\end{definition}
From now on, I will abuse notation, and use ‘$\langle$’ and ‘$\rangle$’ to delimit all ordered sequences, not only pairs. But, speaking strictly, when in the following I denote an ordered triple by $\langle x,y,z\rangle$, that expression should be interpreted as $⟪x, y, z⟫= \langle\langle x,y\rangle,z\rangle$. It doesn't really matter \emph{how} we define ordered tuples; all that matters is that they are well-defined set-theoretic entities which behave as expected. (So, strictly speaking, we'd need to prove a theorem analogous to Theorem \ref{cridorpa} showing that ordered $n$-tuples as just defined behave in the right way. We are not going to do that.)

One remaining case deserves attention. Our notation covers all ordered sequences of two or more members. What about the degenerate case of an ordered sequence with just one member? We shall adopt the following harmless convention (which will however be of use in §\ref{semltwo}): \begin{definition}[Ordered 1-tuple]\label{ordonetup}
	An ordered 1-tuple $\langle x\rangle$ is simply $x$ itself.
\end{definition} This choice has the odd but harmless consequence that everything is a sequence.

\section{Relations}
Intuitively, a relation $R$ characterises the nature of the connections between some things (its \emph{relata}), or an aspect of the structure of some collection of things. We will model a relation set theoretically as a set of \emph{sequences} of entities; we care not just about which things are involved in the relation, but which way they are related. (We need to be able to distinguish when Alice loves Bob from the case where Bob loves Alice, and simply saying that Alice and Bob are related by the `– loves –' relation doesn't allow us to do that.)  So the set $$\{\langle \text{Melbourne}, \text{Victoria}\rangle, \langle \text{Adelaide}, \text{South Australia}\rangle,\ldots\}$$ is (part of) the relation of ` –  is the capital city of  – '.
\begin{definition}[Binary Relation]
	$X$ is a binary relation on a \emph{domain} $D$ iff it is any set that contains nothing except ordered pairs $\langle x,y\rangle$ where $x \in D$ and $y\in D$. ($\emptyset$ contains nothing at all, and so nothing other than ordered pairs, so is a binary relation).
\end{definition}
\begin{definition}[Cartesian Product]
	The \emph{Cartesian product} of sets $X$ and 	$Y$, written‘$X \times Y$’, is the set of all pairs $\langle x,y\rangle$ such that $x \in X$ and $y \in Y$.
\end{definition} A binary relation on $D$ is a subset of $D \times D$. A relation can be generalised to arbitrary numbers of relata: 
\begin{definition}[$n$-place Relation]
  $X$ is an $n$-place relation on $D$ iff it is any set of ordered $n$-tuples of members of $D$. An $n$-place relation is a subset of $\underbrace{D \times \ldots \times D}_{n}$.
\end{definition}
This identification of relations with sets is not necessarily a `metaphysical' hypothesis. Just as in the case of ordered pairs, we need not literally assert that relations are and always were sets. Rather, this is a model that can be used to represent some interesting features of relations. I say some more about the metaphysics of relations in \autoref{c:lequals}.


\paragraph{Properties of Binary Relations} Binary relations have many interesting properties. We can often represent these properties \emph{graphically}, if the domain on which the relation is defined is finite. 
\begin{definition}[Finite Directed Graph]
  A \emph{finite directed graph} on a domain $D$ is a pair $\langle V, A\rangle$ where the set of \emph{vertices} $V$ is $D$, $D$ is finite, and $A$ is a set of \emph{arrows}, directed lines from members of $V$ to members of $V$. (Not every member of $V$ needs to have an arrow attached to it.)
\end{definition} A graph on $D$ represents a binary relation $R$ on $D$ just in case \begin{enumerate}
  \item $\langle x,y\rangle \in R$ iff $x\in V$ and $y \in V$.
  \item $\langle x,y\rangle \in R$ iff there is an arrow from $x$ to $y$ in $A$.
\end{enumerate}
\paragraph{Reflexivity, Transitivity, and Symmetry}
Consider the domain $D=\{1,2,3,4\}$.
\begin{definition}[Reflexive]
  A relation on $D$ is \emph{reflexive} iff for all $x\in D$, $\langle x,x\rangle\in R$. 
\end{definition} Graphically, that means, as in the figure on the left in Figure \ref{fone}, that each vertex has an arrow pointing to itself. A relation is \emph{irreflexive} iff no vertex has an arrow pointing to itself.
\begin{definition}[Transitive]
  A relation on $D$ is \emph{transitive} iff for all $x,y,z\in D$, if both $\langle x,y\rangle\in R$ and $\langle y,z\rangle\in R$, then also $\langle x,z\rangle\in R$.
\end{definition} Graphically, that means, as in the figure on the right in Figure \ref{fone}, that whenever there is an indirect sequence of arrows between two vertices, there is also a `shortcut'. A relation is \emph{intransitive} iff there is never a shortcut.
\begin{figure}
\begin{center}
  {~\xymatrix{1 \ar@(ur,dr)[lr]  & 2 \ar@(ur,dr)[lr] \\
  3 \ar@(ur,dr)[lr] & 4 \ar@(ur,dr)[lr] }\qquad\qquad}{\xymatrix{1  & 2 \ar[l] \\ 3 \ar[ur]
    \ar[u] & 4 \ar[l] \ar[ul] \ar[u] 
    }}
\end{center}  \caption{Reflexivity and Transitivity\label{fone}}
\end{figure}
\begin{definition}[Symmetric]
  A relation $R$ on $D$ is \emph{symmetric} iff whenever $\langle x,y\rangle \in R$, then $\langle y,x\rangle \in R$.
\end{definition} Graphically, this means (as on the left in Figure \ref{ftwo}) that whenever there is an arrow from one vertex to another, there will be a return arrow. A relation is \emph{asymmetric} iff there are no returning arrows at all.
\begin{definition}[Anti-symmetric]
  A relation $R$ on $D$ is \emph{anti-symmetric} iff whenever $\langle x,y\rangle \in R$ and $\langle y,x\rangle \in R$, then $x=y$ (sometimes this is known as \emph{weak asymmetry}).
\end{definition} Graphically, this means (as on the right in Figure \ref{ftwo}) that the only `returning' arrows are loops.
\begin{figure}
\begin{center}
  ~{\xymatrix{1 \ar@/^/[r]  & 2 \ar@/^/[l] \ar@/^/[d] \\
  3 \ar@/^/[r] & 4 \ar@/^/[l] \ar@/^/[u] }\qquad\qquad}{\xymatrix{1 \ar@(ul,dl)[lr]  & 2 \ar[l] \\ 3 \ar[ur]
    \ar[u] & 4 \ar@(dl,dr)[lr] \ar[l] \ar[ul] \ar[u] 
    }}
\end{center}  \caption{Symmetry\label{ftwo}}
\end{figure}
We need to clearly distinguish non-reflexivity from irreflexivity: the former is just the failure of reflexivity – i.e., at least one object in the domain doesn't bear $R$ to itself – while the latter involves every object in the domain not bearing $R$ to itself. Likewise, we need to distinguish non-transitivity from intransitivity, and non-symmetry from asymmetry \emph{and} antisymmetry. 

The observant reader will have noticed that, while we defined \emph{reflexive on $D$}, \emph{symmetric on $D$}, etc., the only definition which actually mentions $D$ is the definition of reflexivity. All the other definitions are \emph{conditional} in form: they say, if certain pairs are in the relation, then certain other pairs will be too. We needn't specify a domain to check whether these conditionals hold of any relation. But to check whether a relation is reflexive, we need not only the ordered pairs of the relation, but also what the domain is, so we can see if any members of the domain are missing from the relation. Reflexivity is an extrinsic property of a relation, because it is relative to the domain from which the relata of the relation may be drawn. The others are intrinsic properties of a relation; whether the relation has them is fixed by which pairs are in the relation. The very same set of pairs might be reflexive on one domain and non-reflexive on another, but it will be symmetric on every domain if it is symmetric on any.

\paragraph{Equivalence Relations}
\begin{definition}[Equivalence Relation]
  If $R$ on $D$ is reflexive, transitive, and symmetric, then it is an \emph{equivalence relation}. 
\end{definition}
An equivalence relation divides up the domain into \emph{equivalence classes}, every member of which bears the relation to every other member of it. The relation `is the same height as' is an equivalence relation on the domain of all people; we say that it \emph{induces a partition} on the domain of people, sorting them into groups which are uniform with respect to height.

The most obvious equivalence relation on a domain $D$ is the identity relation, the set of all pairs $\langle x, x\rangle$ such that $x\ in D$. It is reflexive by definition, and moreover symmetric and transitive; since there are \emph{never} arrows from any $x$ to any distinct $y$, it follows that every time there are such arrows, there are return arrows and shortcuts. (Notice that the identity relation is both symmetric and anti-symmetric.)

Any relation on the empty set is also trivially an equivalence relation. While every empty relation is trivially symmetric and transitive, it will generally fail to be reflexive unless the domain is empty too.

\paragraph{Seriality}
Reflexivity guarantees that every thing in the domain bears the relation to at least one thing, and moreover that among those things is the object itself. If a relation satisfies the first condition, though perhaps without meeting the second, it is \emph{serial}:
\begin{definition}[Serial]
  A relation on $D$ is \emph{serial} iff for all $x\in D$, there is some $y\in D$ such that $\langle x,y\rangle\in R$. 
\end{definition} Graphically, that means, that each vertex has an arrow emerging from it. A relation is \emph{non-serial} if at least one vertex has no arrow emerging from it. (The notion of an `\emph{aserial}' relation – one where no vertex bears the relation to anything – is trivial, since only the empty relation could be aserial.) It is obvious that a reflexive relation is serial. 
\begin{theorem}
A transitive, symmetric, and serial relation is reflexive (and hence is an equivalence relation).
\begin{proof}
  Suppose $R$ is transitive, symmetric, and serial on $D$. Pick an arbitrary member of $D$, $x$. By seriality, there is a $y\in D$ such that $\langle x,y\rangle\in R$. By symmetry, $\langle y,x\rangle\in R$. By transitivity, $\langle x,x\rangle\in R$. Since $x$ was arbitrary, $R$ is reflexive.\end{proof}
\end{theorem}

\paragraph{Connectedness}
An equivalence relation divides up the domain into isolated groups of mutually interrelated elements. A connected relation, intuitively, is one where no item is isolated. But there are a number of things we could mean by this. First, we define a (directed) path: \begin{definition}[Path]
  There is a \emph{path} from $x$ to $y$ in $R$ iff there exists a sequence $z_{1},\ldots,z_{n}$ such that $\langle x,z_{1}\rangle \in R$ and
  $\langle z_{1},z_{2}\rangle \in R$ and \ldots and $\langle z_{n},y\rangle \in R$.
\end{definition}
Now we may define: \begin{definition}[Strong Connectedness]
  $R$ on $D$ is \emph{connected} iff for any $x \in D$ and $y \in D$, such that $x$ and $y$ are distinct, then there is a path from $x$ to $y$ in $R$. 
 \end{definition} This is illustrated in the middle graph in \autoref{fconn}, where by following the arrows one can get from any vertex to any other, perhaps by a number of intermediate vertices.

 There is another property of binary relations, sometimes called \emph{weak connectedness}, which holds iff for any nodes on the graph of $R$, $x$ and $y$, there is some path that joins $x$ to $y$ moving only along arrows (perhaps in the `backwards' direction). This is illustrated in the left hand graph in \autoref{fconn}. For any binary relation $S$, let $S^{*}$ be the relation defined by $\langle x,y\rangle \in S^{*}$ iff $\langle x,y\rangle \in S$ or $\langle y,x\rangle \in S$. 
 \begin{definition}[Weak Connectedness]
  A relation $R$ is weakly connected on $D$ iff  $R^{*}$ is connected on $D$.
\end{definition}
There is an even stronger condition than connectedness, illustrated in the right hand graph in \autoref{fconn}: \begin{definition}[Totality]
  A relation $R$ on $D$ is \emph{total} iff for any $x$ and $y$ in $D$, either $\langle x,y\rangle \in R$ or $\langle y,x\rangle \in R$.
\end{definition}
\begin{figure}
\begin{center}
  ~{\xymatrix{1 \ar[r]  & 2  \ar[d] \\
  3  & 4 \ar[l]}}\qquad\qquad
  {\xymatrix{1 \ar[r]  & 2  \ar[d] \\
  3 \ar[u]  & 4 \ar[l]}}\qquad\qquad
    {\xymatrix{1 \ar@(ul,dl)[lr]  & 2 \ar@(ur,dr)[lr]\ar[l] \\ 3 \ar@(dl,dr)[lr] \ar[ur]
    \ar[u] & 4 \ar@(dl,dr)[lr] \ar[l] \ar[ul] \ar[u] 
    }}
\end{center}  \caption{Connectedness and Totality\label{fconn}}
\end{figure}\begin{theorem}\label{tot}
  If $R$ is total on $D$, then $R$ is reflexive on $D$.
  \begin{proof}
     If $R$ is total, then for any $x$ and $y$ in $D$, either $\langle x,y\rangle \in R$ or $\langle y,x\rangle \in R$. Suppose $x=y$; then either either $\langle x,x\rangle \in R$ or $\langle x,x\rangle \in R$. Therefore, for any $x\in D$, $\langle x,x\rangle \in R$, i.e., $R$ is reflexive on $D$.
   \end{proof} 
\end{theorem}


\paragraph{Inverse and Complement Relations}
Given a relation $R$ on $D$, we can define two interesting further relations in terms of it: \begin{definition}[Inverse] \label{definverse}
Given a relation $R$, its \emph{inverse} $R^{-1}$ is defined to be this set: $\left\{\langle y,x\rangle:\langle x,y\rangle \in R\right\}$. \end{definition}
\begin{definition}[Complement]
Given a relation $R$ on $D$, its \emph{complement} $R'$ on $D$ is defined to be this set: $(D \times D) \setminus R$ – i.e., the set of all ordered pairs of elements of the domain \emph{not} in $R$.
 \end{definition} Note that a complement relation is always with respect to the underlying domain – there is no such thing as \emph{the} complement of a relation – whereas we can define an inverse relation just with reference to the original relation. 

\begin{theorem}[Property Preservation]
  \begin{itemize}
    \item If $R$ is reflexive on $D$ then $R^{-1}$ is reflexive on $D$ and $R'$ is irreflexive on $D$;
    \item  if $R$ is symmetric then $R^{-1}$ and $R'$ are symmetric; 
    \item if $R$ is transitive then $R^{-1}$ is transitive;
    \item  if $R$ is connected on $D$ then $R^{-1}$ is connected on $D$.
  \end{itemize}
\begin{proof}
  I prove one case: reflexivity. If $R$ is reflexive on $D$, then for all $x \in D$, $\langle x,x\rangle \in R$. By definition of inverse, then, each $\langle x,x\rangle \in R^{-1}$; i.e., $R^{-1}$ is also reflexive on $D$.  By definition of complement, \emph{no} $\langle x,x\rangle \in R'$; so $R'$ is irreflexive on $D$.
\end{proof}\end{theorem}



\paragraph{Orderings}
\begin{definition}[Ordering]\label{def:order}
  A relation $R$ on $D$ is a \begin{description}
    \item [Partial order] iff it is reflexive, transitive, and anti-symmetric (e.g., $\subseteq$ on a set of sets; see Figure \ref{fthree});
    \item [Strict Partial Order] iff it is transitive and asymmetric (i.e., $\subset$);
    \item [Total Order] iff it is a total partial ordering (i.e., $\leqslant$ on $\mathbb{N}$). Also sometimes known as a \emph{linear order};
    \item [Strict Total Order] iff it is a total strict partial order (i.e., $<$).
  \end{description} 
\end{definition}

\begin{figure}
  \centering ~{\xymatrix{ \{A,B\} \ar@(ul,dl)[]_{\subseteq}   &
  \{A\}\ar[l]^{\subseteq} \ar@(dr,ru)[]_{\subseteq}\\
   \{B\}
  \ar[u]_{\subseteq} \ar@(ul,dl)[]_{\subseteq}
  & \emptyset \ar@(dr,ru)[]_{\subseteq} \ar[u]^{\subseteq} \ar[l]_{\subseteq}
  \ar[ul]_{\subseteq} 
  }
  }\caption{Partial ordering of a set of sets by $\subseteq$\label{fthree}}
\end{figure}

As defined above, a total order is reflexive, transitive, antisymmetric, and total. By Theorem \ref{tot}, a total order is reflexive. So we could have defined a total order as one that is total, transitive, and antisymmetric.

\paragraph{Well-Orderings}

Suppose $R$ is a total order on $D$. If $\langle x,y\rangle\in R$, then we say $x$ \emph{precedes} $y$. If $\langle x,y\rangle \in R$ and $x\neq y$ and there does not exist a $z$ such that $\langle x,z\rangle \in R$ and $\langle z,y\rangle \in R$, then $x$ \emph{immediately precedes} $y$. (We can give exactly similar definitions for \emph{succeeds}.)

An element $x$ of $D$ is \emph{minimal} iff there is no distinct $y$ that precedes $x$. An element $x$ is \emph{least} if it precedes \emph{every distinct} $y \in D$. (Corresponding definitions can be given for maximality and greatest.)

\begin{definition}[Well-ordering]\label{def:wellord}
  $R$ is a well-ordering of $D$ iff it is a total order and every non-empty subset $d$ of $D$ has a least member under $R$.
\end{definition} Example: $\leqslant$ is a well-ordering of the natural numbers $\mathbb{N}$. Here is another well-ordering of $\mathbb{N}$: \begin{equation}
  x \preccurlyeq y \quad\text{ iff }\quad \begin{cases}
    \text{$x$ is even and $y$ is odd; or}\\
    \text{$x$ and $y$ are even and $x\leqslant y$; or}\\
    \text{$x$ and $y$ are odd and $x\leqslant y$}.\\
  \end{cases}
\end{equation} The order induced by $\preccurlyeq$ yields $0, 2, 4, \ldots, 1, 3, 5, \ldots$. In this ordering, every subset of $\mathbb{N}$ has a $\preccurlyeq$-least element. Notice, however, that $1$ is not least, but it has no immediate predecessor. 

The less-than-or-equal-to relation on natural numbers $\leqslant$ is a well-ordering of $\mathbb{N}$. It can be extended in an obvious way to the less-than-or-equal-to relation on the positive and negative integers $\mathbb{Z}$, which we denote $\leqslant_{\mathbb{Z}}$. (This is strictly speaking a different relation, since $\langle -1,0\rangle\in \leqslant_{\mathbb{Z}}$ but $\langle -1,0\rangle\notin \leqslant$.) $\leqslant_{\mathbb{Z}}$ is a total or linear order on $\mathbb{Z}$, but is not a well-ordering, because the set of all negative integers is a non-empty subset of $\mathbb{Z}$ but has no least member under $\leqslant_{\mathbb{Z}}$ (it does have a greatest member though, unlike $\mathbb{N}$). But $\mathbb{Z}$ can be well-ordered: $x \leqslant_{\text{abs}} y$ iff either (i) $|x|\leqslant |y|$ or (ii) $|x|=|y|$ and $x\leqslant y$, where `$|x|$' here denotes the absolute value of $x$.  This induces the order $0,-1,1,-2,2,\ldots$.

\section{Functions}

Consider the relation `– is the capital city of –', conceived of as relating  Australian states to Australian cities. This relation has an interesting feature: if a state stands in the relation to a city, then it stands in the relation to exactly one city. This is in fact made explicit by the way we have specified the relation: a city C stands in the relation to a state S iff C is (is identical to) the capital city of S, where the descriptive noun phrase `the capital city of S' manages to successfully denote something just in case there is only one capital city. Not all relations have this feature: `$x$ is taller than $y$', for example, cannot be formulated as an identity between $x$ and something denoted by a descriptive phrase involving $y$.

Relations which have this special feature, where whenever $xRy$, then there is no distinct $z$ such that $xRz$, are known as \emph{functions}. 
\begin{definition}[Unary Function]
  If $R$ is a binary relation such that for any $x,y,z$, if $\langle x,y\rangle\in R$ then $\langle x,z\rangle\in R$ iff $y$ is identical to $z$, then $R$ is a unary (one-place) function. 
\end{definition}
We write $f_{R}(x) =y$ iff $\langle x,y\rangle\in R$. The notation `$f_{R}(x)$' is a noun phrase, typically corresponding to a definite description like `the capital city of $x$'. We say $x$ is the \emph{argument} given to $f$, and $y$ the \emph{value} of $f$ given that argument.  The fact that, for a given argument, a function has a unique value, justifies our talking of `\emph{the} value of the function for a given argument', and justifies the link to natural language definite descriptions as corresponding to a function with an argument – see the discussion of definite descriptions in \autoref{c:edr}. 

The notion of a function can be generalised to an arbitrary $n$-place relation, which will give rise potentially to an $n-1$-place function.
\begin{definition}[Function]
	$f$ is an $n$-place function iff (i) it is a relation (ii) and for any sequence $x_{1},\ldots,x_{n}$ of objects, and any $y,z$, if $\langle x_{1},\ldots,x_{n},y\rangle \in f$ and $\langle x_{1},\ldots,x_{n},z\rangle \in f$, then $y=z$ (i.e., each sequence of things stands in the relation to at most one thing). 
\end{definition}

Since a function is a relation, some of its properties are defined only relative to a domain. There is another notion, also confusingly called the domain of a function, which can be thought of as the things the function accepts as arguments.
\begin{definition}[Domain]
The \emph{domain} (note: different to the domain of a relation) of a function  $f$ is the set $$\mathcal{D}(f)=\{x: \text{there is a $y$ such that } \langle x,y \rangle \in f\}.$$
\end{definition} There is also the range of the function, which are the things the function can yield as values.
\begin{definition}[Range]
	The \emph{range} of $f$ is the set $$\mathcal{R}(f)=\{y: \text{there is a $x$ such that } \langle x,y \rangle \in f\}.$$
\end{definition}
\begin{definition}[Partial and Total]
	A function $f$ is \emph{partial}  on domain $D$ if $\mathcal{D}(f) \subset D$; it is \emph{total} if $\mathcal{D}(f) =D$.
\end{definition} Obviously any function $f$ is total on $\mathscr{D}(f)$, but it may be total on some domains and partial on others. In this sense, totality and partiality of a function is another extrinsic property of the underlying relation.

\paragraph{Operators}
\begin{definition}[Operator]\label{operator}
	If $f$ is a function, and $\mathcal{R}(f) \subseteq \mathcal{D}(f)$, then $f$ is an \emph{operator}.
\end{definition}
Consider the numerical function $\mathsf{sq} = \left\{\langle x,y\rangle: x^{2} = y\right\}$, otherwise written $\mathsf{sq}(x) = x^2$. This generalises in the obvious way to binary functions: if the values of the function are appropriate arguments, it is an operator. An example is the addition function on the domain of integers, $\mathsf{plus} = \left\{\langle x, y, z\rangle : z = x + y\right\}$. This is an arithmetical operator.

Some properties of functions are conveniently illustrated by $\mathsf{plus}$.
\begin{definition}[Commutative]
	A binary function $f$ is commutative iff $$f(x,y) = f(y,x).$$  
\end{definition}
\begin{definition}[Associative]
	A binary operator $f$ is associative iff $$f(x,f(y,z)) = f(f(x,y),z).$$ (We need $f$ to be an operator to ensure that it is defined when its values are treated as arguments in this way.)
\end{definition}
It is obvious that $\mathsf{plus}$ is both commutative and associative: $x+y=y+x$ and $x+(y+z)=(x+y)+z$. A commutative but not associative operator is $\mathsf{mean}(x,y) = \tfrac{x+y}{2}$. Clearly $\mathsf{mean}(x,y)=\mathsf{mean}(y,x)$. (This follows from the commutativity of $\mathsf{plus}$.) But $\mathsf{mean}(x,\mathsf{mean}(y,z))$ is not in general equal to $\mathsf{mean}(\mathsf{mean}(x,y),z)$. An associative but non-commutative operator is the concatenation operator on strings $\conc$, which takes two strings and joins them into a longer string by appending the second to the first. Clearly $(x\conc y)\conc z = x \conc (y\conc z)$: `te'$\conc$`a' = `tea' = `t'$\conc$`ea'. But $x\conc y$ isn't necessarily $y\conc x$: `tea' isn't `ate'.



\paragraph{Properties of functions}

Suppose $f$ is a function, and $X$ and $Y$ are some sets. There are some relational properties that $f$ might have relative to $X$ and $Y$.
 \begin{definition}[Into, Onto, etc.]\label{intoonto}
\begin{description}
	\item [Into] $f$ is a function from $X$ \emph{into} $Y$ iff $\mathcal{D}(f) = X$ and $\mathcal{R}(f)\subseteq Y$.
	\item [Onto] $f$ is a function from $X$ \emph{onto} $Y$ iff $\mathcal{D}(f) = X$ and $\mathcal{R}(f) = Y$. An \emph{onto} function is also known as a \emph{surjection}.
	\item [One-One] $f$ is a \emph{one-one} function from $X$ to $Y$ iff $f$ is into and whenever $x\neq y$, $f(x)\neq f(y)$. A one-one function is also known as an \emph{injection}.
	\item [Bijection] $f$ is a \emph{bijection} from $X$ to $Y$ (or, is a \emph{one-one correspondence}) iff $f$ is one-one and onto.
\end{description}
\end{definition} It is important to remember as always that these features are always relative to the sets $X$ and $Y$. A function that is a bijection between $X$ and $Y$ needn't be a bijection between $X'$ and $Y'$. These properties are illustrated in \autoref{fintoonto}.
\begin{figure}
\centering
\begin{tabular}{cccc}


   {\xy
\xymatrix"*"{%
• \\
• \\
• }%
\POS*\frm<40pt>{.}
\POS-(15,4)
\xymatrix{%
• \ar["*"]  \\
• \ar["*"d]  \\
• \ar["*"] }%
\POS*\frm<40pt>{.}
\endxy} & 
~{\xy
\xymatrix"*"{%
• \\
• \\
• }%
\POS*\frm<40pt>{.}
\POS-(15,4)
\xymatrix{%
• \ar["*"]  \\
• \ar["*"]  \\
• \ar["*"]  \\
• \ar["*"u] }%
\POS*\frm<40pt>{.}
\endxy} & 
~{\xy
\xymatrix"*"{%
• \\
• \\
• \\
• }%
\POS*\frm<40pt>{.}
\POS-(15,4)
\xymatrix{%
• \ar["*"]  \\
• \ar["*"]  \\
• \ar["*"] }%
\POS*\frm<40pt>{.}
\endxy} & 
~{\xy
\xymatrix"*"{%
• \\
• \\
• }%
\POS*\frm<40pt>{.}
\POS-(15,4)
\xymatrix{%
• \ar["*"]  \\
• \ar["*"]  \\
• \ar["*"] }%
\POS*\frm<40pt>{.}
\endxy} \\
Into & Onto & One-one & Bijection
\end{tabular}
  \caption{Properties of functions \label{fintoonto}}
\end{figure}


Some examples; these are all total functions from $\mathbb{N}$: \begin{itemize}
	\item The constant function $f(x) = 1$ is into $\mathbb{N}$.
	\item The function $g(x) = \begin{cases}
		x/2 &\text{ if $x$ is even}\\ (x+1)/2 &\text{ if $x$ is odd} \end{cases}$ is onto $\mathbb{N}$. (Every natural number is the value of this function for two arguments.)
	\item The function $h(x)=x^3$ is one-one from  $\mathbb{N}$ to $\mathbb{N}$. (Every natural number has a unique cube, every cube has a unique cube root – but not every natural number is a cube.)
 	\item The identity function $i(x) = x$ is a bijection from  $\mathbb{N}$ to $\mathbb{N}$.
\end{itemize}

Because a function is a relation, it has an inverse and complemenent. The complement is not very interesting from the perspective of functions, since only in trivial cases will the complement of a function be a function. But the inverse of a function may be of more interest. The definition of an inverse relation in \autoref{definverse} suffices to define an inverse function, but here is an equivalent definition directly in terms of functions: 
\begin{definition}[Inverse] If $g$ is a function, it is the \emph{inverse} of $f$ iff  $\mathcal{D}(f)=\mathcal{R}(g)$, $\mathcal{R}(f)=\mathcal{D}(g)$, and for all $x$, $g(f(x))=x$. In that case, we often write $f^{-1}$ for $g$.
 \end{definition} Obviously, $f$ is also the inverse of $f^{-1}$. 

Note that not every function has an inverse function (it always has an inverse relation, which may or may not be a function):
\begin{theorem}
	A function has an inverse iff it is one-one. \begin{proof}
    Left for exercises.
  \end{proof}
\end{theorem}
\begin{theorem}
If $f$ is one-one but not a bijection, $f^{-1}$ is a partial function. 
\begin{proof}
If $f$ is one-one from $X$ to $Y$ but not a bijection, then there exists a $y\in Y$ which is not the value of $f$ for any member of $X$. Assume for \emph{reductio} that $f^{-1}$ is total on $Y$. Then there must be some $x \in X$ such that $f^{-1}(y) = x$. But then by the definition of inverse, $f(x) = f(f^{-1}(y)) = y$. But then $x$ is a member of $X$ which yields $y$ as value when given to $f$ as argument; contradiction. So $f^{-1}$ is not total.
\end{proof}
\end{theorem}


\section{Size}

\begin{definition}[Finite]
  A set $X$ is finite iff there exists a set of natural numbers $N = \{1,\ldots,n\}$ and a bijection $f$ from $X$ to $N$.
\end{definition} A finite set can be put into one-one correspondence with some initial fragment of the sequence of natural numbers. (This corresponds to the notion, familiar from childhood, of counting some things by pointing to each of them while reciting some initial fragment of the natural numbers.) A set is infinite iff it is not finite. Where there is a bijection between $X$ and $\{1,\ldots,n\}$, $X$ has $n$ members – we say that $X$ has cardinality $n$, $|X|=n$.

Now we will define some further notions useful for talking about \emph{size}.
\begin{definition}[Enumeration] \label{enumer} $f$ is an \emph{enumeration} of $X$ iff $\mathcal{D}(f) \subseteq \mathbb{N}$ (the natural numbers), $\mathcal{R}(f) = X$, and $f$ is onto.
\end{definition} An enumeration associates each member of $X$ with some number. This is a generalisation of the ordinary notion of counting, however: since we only require that an enumeration be onto, all sorts of weird `counting' procedures are included: e.g, the constant function from $\mathbb{N}$ onto $\{a\}$ is an enumeration of $\{a\}$. But this definition does ensure that any enumerable set is intuitively countable: since any enumerable set $X$ is going to be no larger than some subset of $\mathbb{N}$, by \autoref{intoonto}. This is captured in the following definition: 
\begin{definition}[{[Un]}Countable] $X$ is \emph{countable} iff there exists a function which enumerates it; it is \emph{denumerable} iff it is countable and infinite; it is \emph{uncountable} if it is not countable.
\end{definition}
Obviously, all finite sets are enumerable, since if $f$ is a bijection showing $X$ to be finite, $f^{-1}$ exists and is an enumeration of $X$. But some infinite sets are countable too: obviously, $\mathbb{N}$ itself (there is a no bijection between $\mathbb{N}$ and some initial subset of $\mathbb{N}$, but the identity function enumerates it).

The notion of cardinality can be extended to infinite sets too \citep[chs. 3, 6]{macsetthl}. \begin{definition}[Equinumerosity]
	$X$ and $Y$ are \emph{equinumerous} iff there exists a bijection from $X$ to $Y$.
\end{definition}
\begin{definition}[Cardinality]\label{card}
	For each set, $X$ – whether finite or infinite – there exists its cardinality, $|X|$, meeting the following condition: for any $X$ and $Y$, $|X|=|Y|$ iff $X$ and $Y$ are equinumerous.
\end{definition}
Certainly treating the natural numbers as cardinals of finite sets meets the condition in Definition \ref{card}. But Definition \ref{card} applies to any sets; it's just that, for the infinite sets, we have no pre-theoretical grasp on what their cardinalities might be. That's okay though: we can characterise them. \begin{definition}[$\leq$]\label{leq}
	If $\lambda$ and $\mu$ are cardinals, such that $|X|=\lambda$ and $|Y|=\mu$, then $\lambda \leq \mu$ iff there is an injection from $X$ to $Y$. 
\end{definition}
\begin{theorem}
	Let $|Y|=\mu$. Then $\lambda\leq\mu$ iff there exists an $X$ such that $X \subseteq Y$ and $|X|=\lambda$. \begin{proof}
		Obvious: by Definition \ref{leq}, $\lambda\leq\mu$ iff there is an injection from $X$ to $Y$ iff there is a bijection from $X$ to a subset of $Y$.\label{cardleq}
	\end{proof}
\end{theorem}
\begin{theorem}[Schröder-Bernstein]
	$\leq$ partially orders the cardinals \citep[38–41]{macsetthl}. 
\end{theorem}

With the notion of a cardinal, we can prove that there are more sizes of set than finite sizes, because there are more cardinals than just the natural numbers, which are the cardinalities of finite sets. Since there is no bijection from any finite subset of natural numbers to the whole set of natural numbers, the cardinality of the set of natural numbers is not any finite natural number. We introduce $\aleph_{0}$ to name $|\mathbb{N}|$. 
\begin{theorem}[Cantor]\label{cantor}
For any set $X$, $|X| < |\wp(X)|$.
\begin{proof}
	Define a function $f$ such that for all $x \in X$, $f(x) = \{x\}$. Since if $x \in X$, $\{x\} \subseteq X$, $f$ is an injection from $X$ into $\wp(X)$, and hence $|X|\leq|\wp(X)|$ by Definition \ref{leq}.
	
	To show that $|X| < |\wp(X)|$, we need to show that $|X| \neq |\wp(X)|$, i.e. (by Definition \ref{card}), that $X$ and $\wp({X})$ are not equinumerous. Let $g$ be any function from $X$ to $\wp(X)$. The domain of $g$ is the members of $X$, and the range are subsets of $X$, precisely the kinds of things members of $X$ can be members of. But not every member of $X$ needs to be a member of the subset of $X$ picked out by $g(x)$. Let us define the set of such things, the $x\in X$ which aren't members of $g(x)$: $$D = \{x \in X : x \notin g(x)\}.$$
	    Since $D$ consists solely of members of $X$, $D \subseteq X$ and hence $D \in \wp(X)$. If $g$ is a bijection, then there must be some $d \in X$ such that $g(d) = D$. Obviously, then $d \in D$ iff $d \in g(d)$. But by definition of $D$, $d \in D$ iff $d \notin g(d)$. So if $g$ is a bijection, there is some $d$ such that $d \in g(d)$ iff $d \notin g(d)$, and since there obviously is no such $d$, then $g$ cannot be a bijection. Since $g$ was arbitrary, \emph{no} function from $X$ to $\wp(X)$ is a bijection, and therefore $X$ and $\wp(X)$ are not equinumerous.
  \end{proof}
\end{theorem} 
Cantor's theorem shows us that the powerset of a set is always a strictly larger cardinality than the set. We already know that there is one countable infinite set, $\mathbb{N}$. Are there uncountable sets? Yes, obviously: $\wp(\mathbb{N})$: the set of all sets of natural numbers. This allows us to show that another set is uncountable: the real numbers, numbers with arbitrarily many decimal places. Indeed, we can show the following, from which it follows trivially that the set of reals is uncountable. 
\begin{theorem}[Uncountability of the Unit Interval]
The set of real numbers between $0$ and $1$ inclusive, $[0,1]$, is uncountable. \begin{proof}
	(Sketch.) Each subset of $\mathbb{N}$ can be associated with an infinite binary sequence (sequence of $0$s and $1$s): the sequence corresponding to $X \in \wp(\mathbb{N})$ is the one which has a $1$ at position $n$ iff $n \in X$. Each infinite binary sequence corresponds to exactly one real number in $[0,1]$. So there is a bijection from $\wp(\mathbb{N})$ to $[0,1]$, so $|\wp(\mathbb{N})| = |[0,1]|$, and since the former is uncountable, so is the latter.
\end{proof}
\end{theorem}


A final theorem linking countability and set theory, useful – perhaps – for some exercise or other.
\begin{theorem}[Countable Unions]\label{countu}
	If $\mathbf{X} = \{X_{1},\ldots,X_{n},\ldots\}$ is countable, and each $X_{i}$ is countable, then $\bigcup \mathbf{X}$ is countable.
\end{theorem}
I sketch the proof.
Let $P$ be the set of positive prime numbers. This set is in one-one correspondence with $\mathbb{N}$ (for every $n$, there is one and only one $n$-th prime). Since $\mathbf{X}$ is countable, there is a one-one function from $\mathbf{X}$ into the natural numbers, so there is a one-one function $f$ from $\mathbf{X}$ into $P$.

Since each $X_{i}$ is countable, there is a function $g_{i}$ that is a one to one correspondence between $X_{i}$ and a subset of $\mathbb{N}$. Define the function $h_{i}(x) = f(X_{i})^{g_{i}(x)}$. Since every integer has a unique prime decomposition – by the fundamental theorem of arithmetic \citep[§V.14]{fundta} – if $h_{i}\neq h_{j}$ then $\mathcal{R}(h_{i})\cap\mathcal{R}(h_{i})=\emptyset.$

Now define: $h(x) = h_i(x)$ where $x\in X_i$ and $x$ is not a member of any $X_{j}$ where $j < i$. Clearly $\mathcal{D}(h) = \bigcup\mathbf{X}$. Obviously $\mathcal{R}(h) \subseteq \mathbb{N}$. And $h$ is one-one, since each $h_{i}$ is a function and their ranges are disjoint. So $\bigcup\mathbf{X}$ is in one-one correspondence with a subset of $\mathbb{N}$ so is countable. 

{\small
\subsection*{Further Reading}
\addcontentsline{toc}{subsection}{Further Reading}


Another presentation of the set theory needed for this book can be found in
 \citet{bevpospa}. A presentation which goes well beyond what we need, but is philosophically nuanced, is
 \citet{potsetthi}. The iterative conception of a set is discussed by \citet{boolos}.



\subsection*{Exercises} \label{ex:sets}
\addcontentsline{toc}{subsection}{Exercises}

\begin{enumerate}

\item Prove: \begin{enumerate}
	\item If $X \subseteq Y$ and $Y \subseteq X$ then $X=Y$.
	\item If $X \subseteq Y$ and $Y \subseteq Z$ then $X \subseteq Z$.
	\item $X \not\subset X$.
	\item If $X \subset Y$ then $Y \not\subset X$.
	\item If $X \subset Y$ then $X \subseteq Y$.
\end{enumerate}

\item Prove: \begin{enumerate}
	\item $X\cap X=X\cup X=X$.
	\item $X \cap \emptyset = \emptyset$.
	\item $X \cup \emptyset = X$.
	\item $X \subseteq Y$ iff $X \cap Y = X$ iff $X \cup Y = Y$.
	\item $X \cap (Y \cup Z) = (X \cap Y) \cup (X \cap Z)$.
	\item $X\setminus (X \cap Y) = X\setminus Y$.
	\item $X \setminus (Y \cap Z) = (X\setminus Y)\cup(X\setminus Z)$.
\end{enumerate}

\item Prove that if there is a universal set $\Omega$, then the axiom of separation entails a contradiction.

\item Let us define a \emph{Kuratowski ordered pair} $« x,y»$ as the set $\{\{x\},\{x,y\}\}$. Prove that \begin{enumerate} 
	\item $« x,y» = « u,v»$ iff $x=u$ and $y=v$.
	\item If $x=y$ then $« x,y» = \{\{x\}\}$.
\end{enumerate}

\item \begin{enumerate}
	\item Prove that if $X$ has $n$ members, $\wp(X)$ has $2^{n}$ members.
	\item Prove that $X \subseteq Y$ iff $\wp(X) \subseteq \wp(Y)$.
	\item Show, by providing a counterexample, that it is not always true that $\wp(X) \cup \wp(Y) = \wp(X \cup Y)$.
\end{enumerate}


\item Which of the relations expressed by the following English predicates are equivalence relations: \begin{enumerate}
  \item `$x$ and $y$ attend the same lectures', on the domain of Oxford students.
  \item `$x$ is studying the same subject as $y$', on the domain of Oxford students.
\end{enumerate}
\item Prove the following: \begin{enumerate}
\item If $R$ is irreflexive, then $R'$ is reflexive.
  \item If $R$ is symmetric, then $R^{-1}$ is symmetric.
  \item If $R$ is symmetric, then $R'$ is symmetric.
  \item If $R$ is transitive, then $R^{-1}$ is transitive.
  \item If $R$ is connected, then $R^{-1}$ is connected.
  \item If $R$ is asymmetric and non-empty, then $R'$ is non-symmetric.
\end{enumerate}
\item \begin{enumerate}
  \item If $R$ is transitive, is $R'$ transitive?
  \item If $R$ is connected, is $R'$ connected?
  \item If $R$ is antisymmetric, is $R'$ symmetric?
  \item If $R$ is transitive and non-empty, is $R'$ is transitive?
\end{enumerate}
\item If $R$ is defined on the empty domain, 
    can it be reflexive? Can it be irreflexive? What about
    transitivity and symmetry?
\item What is wrong with the following argument that reflexivity is a consequence of symmetry and transitivity? \begin{quote}
  If $\langle x,y\rangle \in R$, then $\langle y,x\rangle\in R$ since we assume $R$ is symmetric. If both $\langle x,y\rangle\in R$ and $\langle y,x\rangle \in R$, then since $R$ is transitive, $\langle x,x\rangle\in R$ – so $R$ is reflexive. (After \citealt[p.\ 52]{pmwmatmel}.)
\end{quote} 
\item \begin{enumerate}
  \item Can a relation be asymmetric and reflexive?
  \item Can a relation 
      be transitive, non-symmetric and irreflexive?
  \item   Can a relation 
        be connected and irreflexive?
\end{enumerate}  
\item  A relation $R$ satisfies \emph{trichotomy} iff, for all $x$ and $y$, at most one of these three holds: $Rxy$, $Ryx$, or $x=y$. Is every trichotomous relation connected? Under what circumstances is a connected relation trichotomous?


\item Show that a well-ordering of $D$ is a total ordering
    of $D$, but not \emph{vice versa}.

  
  



\item Prove that \begin{enumerate}
  \item If $x$ is a least member under an ordering $R$, then it is the unique least member.
  \item The set of natural numbers $\mathbb{N}$ is well-ordered by $<$.
  \item The set of (positive and negative) integers $\mathbb{Z}$ is \emph{not} well-ordered by $<$.
  \item There exists orderings with no maximal elements.
\end{enumerate}

\item A relation $R$ is \emph{dense} iff whenever $\langle x,y\rangle \in R$, there exists a $z$ such that $\langle x,z\rangle \in R$ and $\langle z,y\rangle \in R$. Prove that on the domain of the natural numbers $\mathbb{N}$, the greater-than relation $>$ is not dense; but that on the domain of the positive rationals (i.e., numbers of the form $\frac{n}{m}$ where $n,m \in \mathbb{N}$), it is dense.

\item Let $D = \{1,2,3,5,6,10,15,30\}$. Let $R$ be the relation on $D$ defined by \begin{equation*}
  R = \{\langle x,y\rangle: x\text{ divides $y$ without remainder}\}.
\end{equation*}\begin{enumerate}
  \item Show that $R$ is a weak partial order but not a total order.
  \item Draw the graph of $R$, and identify any minimal, maximal, least or greatest elements.
  \item Do the same for the set $\wp \{a,b,c\}$, using the relation $\subseteq$ on that domain.
\end{enumerate}

\item \begin{enumerate}
\item Show that if $f^{-1}$ is the inverse of $f$, then $f$ is also the inverse of $f^{-1}$.
	\item Show that $f$ has an inverse function iff it is one-one.
	\item Give an example of a function that \begin{enumerate}
		\item is into but is not onto;
		\item is one-one but is not a bijection;
		\item has no inverse;
		\item is its own inverse. (A function which is its own inverse is known as an \emph{involution}.)
	\end{enumerate}
	\item Under exactly what conditions is the union of two functions itself a function? (I.e., state necessary and sufficient conditions.)
	\item Give an example of a set which is \begin{enumerate}
		\item Countable;
		\item Denumerable;
		\item Uncountable. (Hint: use \autoref{cantor} and 6.(d).ii.)
	\end{enumerate} 
\end{enumerate} 



\item \begin{enumerate}
	\item Show that if $X$ is countable, then if $Y \subseteq X$, $Y$ is countable.
\item Show that if $X$ and $Y$ are both countable, $X \times Y$ is countable. (Hint: show that the set of ordered pairs of natural numbers is countable.)

\end{enumerate}





\end{enumerate}

Answers to selected exercises on page \pageref{ans:sets}.
}

